---
title: Correcting batch effects in single-cell RNA-seq data
author: 
- name: Aaron Lun
  affiliation: Cancer Research UK Cambridge Institute, Cambridge, United Kingdom
date: "Revised: 3 February 2019"
output:
  BiocStyle::html_document:
    toc_float: true
package: batchelor
bibliography: ref.bib
vignette: >
  %\VignetteIndexEntry{1. Correcting batch effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}    
---

```{r, echo=FALSE, results="hide", message=FALSE}
require(knitr)
opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r setup, echo=FALSE, message=FALSE}
library(batchelor)
set.seed(100)
```

# Introduction

Batch effects refer to differences between data sets generated at different times or in different laboratories.
These often occur due to uncontrolled variability in experimental factors, e.g., reagent quality, operator skill, atmospheric ozone levels.
The presence of batch effects can interfere with downstream analyses if they are not explicitly modelled.
For example, differential expression analyses typically use a blocking factor to absorb any batch-to-batch differences.

For single-cell RNA sequencing (scRNA-seq) data analyses, explicit modelling of the batch effect is less relevant. 
Manny common downstream procedures for exploratory data analysis are not model-based, including clustering and visualization.
It is more generally useful to have methods that can remove batch effects to create an corrected expression matrix for further analysis. 
This follows the same strategy as, e.g., the `removeBatchEffect()` function in the `r Biocpkg("limma")` package [@ritchie2015limma].

Batch correction methods designed for bulk genomics data usually require knowledge of the other factors of variation.
This is usually not known in scRNA-seq experiments where the aim is to explore unknown heterogeneity in cell populations.
The `r Biocpkg("batchelor")` package implements batch correction methods that do not rely on _a priori_ knowledge about the population structure.
To demonstrate, we will use a small scRNA-seq data set [@tasic2016adult] from the `r Biocpkg("scRNAseq")` package:

```{r}
library(scRNAseq)
data(allen)

library(SingleCellExperiment)
sce1 <- as(allen, "SingleCellExperiment")
counts(sce1) <- assay(sce1)

library(scater)
sce1 <- sce1[1:2000,] # reducing the size for demo purposes.
sce1 <- normalize(sce1) # quick and dirty normalization.
```

We artificially create a batch effect in a separate `SingleCellExperiment` object:

```{r}
sce2 <- sce1
logcounts(sce2) <- logcounts(sce2) + rnorm(nrow(sce2), sd=2)

combined <- cbind(sce1, sce2)
combined$batch <- factor(rep(1:2, c(ncol(sce1), ncol(sce2))))

plotPCA(combined, colour_by="batch") # checking there is a batch effect.
```

# Mutual nearest neighbors

## Overview

Mutual nearest neighbors (MNNs) are defined as pairs of cells - one from each batch - that are within each other's set of `k` nearest neighbors.
The idea is that MNN pairs across batches refer to the same cell type, assuming that the batch effect is orthogonal to the biological subspace [@haghverdi2018batch].
Once MNN pairs are identified, the difference between the paired cells is used to infer the magnitude and direction of the batch effect.
It is then straightforward to remove the batch effect and obtain a set of corrected expression values.

## The new, fast method

The `fastMNN()` function performs a principal components analysis (PCA) to obtain a low-dimensional representation of the input data.
MNN identification and correction is performed in this low-dimensional space, which offers some advantages with respect to speed and denoising.    
This returns a `SingleCellExperiment` containing a matrix of corrected PC scores is returned, which can be used directly for downstream analyses such as clustering and visualization.

```{r}
f.out <- fastMNN(A=sce1, B=sce2)
str(reducedDim(f.out, "corrected"))
```

The batch of origin for each row/cell in the output matrix is also returned:

```{r}
rle(f.out$batch)
```

Another way to call `fastMNN()` is to specify the batch manually.
This will return a corrected matrix with the cells in the same order as that in the input `combined` object.
In this case, it doesn't matter as the two batches were concatenated to created `combined` anyway, 
but these semantics may be useful when cells from the same batch are not contiguous.

```{r}
f.out2 <- fastMNN(combined, batch=combined$batch)
str(reducedDim(f.out2, "corrected"))
```

As we can see, the batch effect is successfully removed in the PCA plot below.

```{r}
plotReducedDim(f.out, colour_by="batch", use_dimred="corrected")
```

We can also obtain per-gene corrected expression values by using the rotation vectors stored in the output.
This reverses the original projection used to obtain the initial low-dimensional representation.
There are, however, `r Biocpkg("simpleSingleCell", "batch.html#using-the-corrected-values-in-downstream-analyses", "many caveats")` to using these values for downstream analyses.

```{r}
cor.exp <- assay(f.out)[1,]
hist(cor.exp, xlab="Corrected expression for gene 1") 
```

While the default arguments are usually satisfactory, there are _many_ options for running `fastMNN()`, e.g., to improve speed or to achieve a particular merge order.
Refer to `r Biocpkg("simpleSingleCell", "batch.html", "the corresponding workflow")` for more details.

## The old, classic method

The original method described by @haghverdi2018batch is implemented in the `mnnCorrect()` method.
This performs the MNN identification and correction in the gene expression space, and uses a different strategy to overcome high-dimensional noise.
`mnnCorrect()` is called with the same semantics as `fastMNN()`:

```{r}
classic.out <- mnnCorrect(sce1, sce2)
```

... but returns the corrected gene expression matrix directly, rather than using a low-dimensional representation^[Again, those readers wanting to use the corrected values for per-gene analyses should consider the `r Biocpkg("simpleSingleCell", "batch.html#using-the-corrected-values-in-downstream-analyses", "caveats")` mentioned previously.].
This is wrapped in a `SummarizedExperiment` object to store various batch-related metadata.

```{r}
classic.out
```

For scRNA-seq data, `fastMNN()` tends to be both faster and better at achieving a satisfactory merge.
`mnnCorrect()` is mainly provided here for posterity's sake, though it is more robust than `fastMNN()` to certain violations of the orthogonality assumptions.

# Batch rescaling

`rescaleBatches()` effectively centers the batches in log-expression space on a per-gene basis.
This is conceptually equivalent to running `removeBatchEffect()` with no covariates other than the batch.
However, `rescaleBatches()` achieves this rescaling by reversing the log-transformation, 
downscaling the counts so that the average of each batch is equal to the smallest value, and then re-transforming.
This preserves sparsity by ensuring that zeroes remain so after correction, and mitigates differences in the variance when dealing with counts of varying size between batches^[Done by downscaling, which increases the shrinkage from the added pseudo-count.].

Calling `rescaleBatches()` returns a corrected matrix of per-gene log-expression values, wrapped in a `SummarizedExperiment` containin batch-related metadata.

```{r}
rescale.out <- rescaleBatches(sce1, sce2)
rescale.out
```

While this method is fast and simple, it makes the strong assumption that the population composition of each batch is the same. 
This is usually not the case for scRNA-seq experiments in real systems that exhibit biological variation.
Thus, `rescaleBatches()` is best suited for merging technical replicates of the same sample, e.g., that have been sequenced separately.

# Other utilities

## Multi-batch normalization

Differences in sequencing depth between batches are an obvious cause for batch-to-batch differences.
These can be removed by `multiBatchNorm()`, which downscales all batches to match the coverage of the least-sequenced batch.
This function returns a list of `SingleCellExperiment` with log-transformed normalized expression values that can be directly used for further correction.

```{r}
normed <- multiBatchNorm(A=sce1, B=sce2)
names(normed)
```

Downscaling mitigates differences in variance between batches due to the mean-variance relationship of count data. 
It is achieved using a median-based estimator to avoid problems with composition biases between batches [@lun2016pooling].
An example usage of this function is provided in `r Biocpkg("simpleSingleCell", "batch.html#feature-selection-across-merges", "this workflow")`.

## Multi-batch PCA

Users can perform a PCA across multiple batches using the `multiBatchPCA()` function.
The output of this function is roughly equivalent to `cbind`ing all batches together and performing PCA on the merged matrix. 
The main difference is that each sample is forced to contribute equally to the identification of the rotation vectors.
This allows small batches with unique subpopulations to contribute meaningfully to the definition of the low-dimensional space.

```{r}
pca.out <- multiBatchPCA(A=sce1, B=sce2)
names(pca.out)
```

This function is used internally in `fastMNN()` but can be explicitly called by the user to perform hierarchical merges - 
see `r Biocpkg("simpleSingleCell", "batch.html#controlling-the-merge-order", "here")` for details.

# Session information

```{r}
sessionInfo()
```

# References
